---
title: "Multivariate Phylogenetic Comparative Methods"
author: "Dean Adams (dcadams@iastate.edu)"
date: "`r Sys.Date()`"
output: 
  html_document:
    css: ../../Instructor_Materials/style/styles.css
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../") })
---



```{r,echo=F}

knitr::read_chunk('MultivariatePCMs.R')

```



### **Overview**
Today we examine how one can perform phylogenetic comparative analyses on multivariate datasets. For this, the phenotypic data (Y) is a N x p matrix of phenotypic values for N species, across p trait dimensions. These p-dimensions could be a set of univariate traits (e.g., length, width, height, etc.) or they could represent a multi-dimensional trait encoded by multiple numbers (e.g., shape from geometric morphometric methods). The goal is to evaluate patterns in the response variables (Y) while conditioning the data on the phylogeny. In other words, we wish to perform macroevolutionary analyses via phylogenetic comparative methods, but do so on multivariate data. 

As we discussed in class, most univariate PCMs now have a multivariate counter-part: phylogenetic regression, phylogenetic ANOVA, phylogenetic correlation, phylogenetic signal, and comparing rates of phenotypic evolution. What does *NOT* have a proper multivariate counter-part are multivariate evolutionary models beyond BM1 and BMM (see Adams and Collyer 2018a; Adams and Collyer 2019). 

### Download data files
Today we will use data from the following files: 

[Phylogeny](https://eeob-macroevolution.github.io/EEOB-565X-Spring2018/practicals/TutorialData/Consensus of 1000 salamander trees.nex)

[Data1](https://eeob-macroevolution.github.io/EEOB-565X-Spring2018/practicals/TutorialData/meandata-CinGlutOnly.csv)

Additionally, The R-script for running the code found in this tutorial is found here:

[RScript](https://eeob-macroevolution.github.io/EEOB-565X-Spring2018/practicals/TutorialData/MultivariatePCMs.r)

As before, on your local computer, place them in a folder named 'TutorialData', which is found in the same directory as this tutorial. 

### **Visualizing High-Dimensional Data**
One challenge with multivariate data is that visualizing patterns in such data is not always straightforward. For instance, we may have 8 phenotypic measurements per species (`TailLength`, `HeadLength`, `Snout.eye`, `BodyWidth`,  `Forelimb`, and  `Hindlimb`), meaning each species is a point in an 8-dimensional phenotype space. Visually characterizing patterns of dispersion in such spaces is challenging, so we rely on ordination methods from classical multivariate statistics. To do this, we first bring our data into R. Then we perform a principal components analysis to visualize a two-dimesional representation (projection) of our data that explains as much variation as is possible such that the dataspace is not distored (i.e., under the constraint of rigid rotations).

The key for much of the principal components analysis that we'll do relies on `gm.prcomp()`, an extension of `prcomp()` that can also account for phylogeny.

<div class = "func">

<details>
  <summary> <font size="+2"> `corHMM()` </font> </summary>
 
Function performs principal components analysis (PCA) or phylogenetically-aligned components (PaCA) on Procrustes shape coordinates.This function has a few different arguments but we'll primarily be concerned with the following:

* $A$ Can be a 3D array but for our purposes it will be multivariate data in an $n\times p$ matrix.
* $phy$ a phylo object
* $GLS$ The value is defaulted to `FALSE` and is a logical for whether GLS-centering and covariance estimating should be used. This is where we use the information from the phylogeny in the analytical computation of our principal components

This function The principal components analysis. Most of the elements that are returned are to be used in other functions for analysis.


</details>
</div>
<br></br>

```{r,read_data, eval=TRUE}

```



#### **Phylomorphospaces**
When one has a phylogeny, one useful visualization is a phylomorphospace. Here, an ordination is performed on the phenotypic data at the tips of the phylogeny, ancestral states are estimated, and these are projected into the dataspace. This forms a plot that contains both the tips, and the phylogeny; allowing a visual inspection of how trait dispersion relative to the phylogenetic relationships among taxa. Such plots are very useful for hypothesis generation, as discussed in class. 

We can plot phylomorphospace similar to our plotting of the PCA, however, we need to add an additional argument `phylo=TRUE` to the `plot()` function.

```{r,Phylomorphospace, eval=TRUE}
 
```

Notice that the points don't move at all between our PCA and phylomorphospace plots, this is because the latter merely plots the phylogeny in the principal components analysis. We are **NOT** accounting for phylogeny in the analytical step of computing our principal components

#### **Phylo PCA**

We can directly account for the phylogeny in our principal components analysis through GLS. Much like GLS can be used in regression to account for phylogeny by using the phylogenetic covariance matrix in the error term, we can use the phylogeny for covariance estimation and centering.

To perform phylo PCA we only need to set `GLS=TRUE` in the `gm.prcomp()` function.

```{r,phylo_pca,eval=TRUE}

```

#### **Phylogenetic Regression**
For univariate response data, the 'workhorse' of PCMs is undoubtedly phylogenetic regression. Three implementations are commonly used and all lead to identical results when implemented properly: phylogenetically independent contrasts (PIC), phylogenetic generalized least squares (PGLS), and phylogenetic-transformation with OLS regression. Unfortunately, for multivariate data, as trait dimensionality increases, the power of these methods decreases, because test statistics are evaluated using parametric approaches (Adams 2014a; Adams and Collyer 2015). 

One solution is to use permutation methods. Specifically, phylogenetic transformation of the X and Y data is performed, and residuals from a reduced model are permuted (RRPP) to obtain significance (Adams and Collyer 2018b). The method has been shown to display appropriate type I error and has high power. It is implemented below.

To accomplish this we use `procD.pgls()`, it works much like `lm()`, `glm()`, or `lm.rrpp()` in that we need to provide a formula for our regression. They main difference is that there is also a `phy` argument and a number of iterations `iter` since we are using permutation. However, before we begin our regression, we need to put our data in a special geomorph dataframe with `geomorph.data.frame()`.

```{r,phylo_reg, eval=TRUE}
 
```



#### **Phylogenetic ANOVA**
Statistically related to regression is analysis of variance (ANOVA). It has recently been shown that for phylogenetic data, the RRPP approach described above is equally appropriate for phylogenetic ANOVA (Adams and Collyer 2018b). Recall however that the dispersion of groups across the phylogeny can affect both statistical and biological inferences. In particular, 'clumping' of groups on the phylogeny (phylogenetic aggregation) can lower statistical power and hamper inferences. In the extreme case of all species for a group belonging to a monophyletic sub-clade, one effectively has a single 'transition' of groups on the phylogeny; making it challenging to identify group differences in the response variable (this is akin to the BiSSE issues of state change replication discussed earlier in the semester). 

```{r,phylo_ANOVA, eval=TRUE}
```

#### **Phylogenetic PLS: Phylogenetic Partial Least Squares**
Tests of association between two traits are commonly accomplished using correlation. The multivariate equivalent is partial least squares (PLS), which identifies the maximal assocation between two sets of variables. This may also be implemented in a phylogenetic context, with significance obtained via RRPP (Adams & Felice 2014).

```{r,phylo_PLS, eval=TRUE}
```

#### **Phylogenetic Signal**
The degree to which trait variation associates with the phylogeny is termed phylogenetic signal. For univariate data, Blomberg et al. (2003) proposed the Kappa statistic. Its multivariate equivalent, K.mult, evaluates phylogenetic signal in multivariate data (Adams 2014b).

```{r,phylo_signal, eval=TRUE}
```

#### **Multivariate Rates of Evolution**
One can also envision comparing rates of phenotypic evolution in multivariate traits. Here, the net rate of evolution is characterized, found as the mean of the rates of the individual trait dimensions (Adams 2014c). Two approaches for comparing evolutionary rates have been developed, which mirror methods discussed for univariate rate tests. First, one may compare rates of multivariate evolution between clades (Adams 2014c). Second, for two multivariate traits, one may compare rates of evolution between traits (Denton and Adams 2015).

```{r,evol_rates, eval=TRUE}
```



<div class = "try">
### Plethodontid landmarks as multivariate data 

Plethodontid salamanders have a lot of diversity in their head shapes. Since quantifying head shape is better done in a multivariate setting, we have took landmark measurements of the head. We're interested in asking whether there is a relationship between the shape of salamander head and head size.

`geomorph` actually has a built in dataset that we can use for this analysis. We can load an format the data set with the following commands:

```{r,pleth_read,eval=FALSE}
```

We now have our salamander tree in the saved as `pleth_tree` and our head landmark data in `landmark`. We then performed a Procrustes analysis on our landmark data to align it. within `procD_landmarks` we can notice `procD_landmarks$coords` contains our aligned landmarks while `procD_landmarks$Csize` is our centroid size, or head size in this case. 


1. Visualize the alligned landmark data with principal components, using both phylomorphospace and phylo PCA. How do these plots compare to one another? 
2. use `geomorph.data.frame()` to format our aligned data set. 
3. Perform multivariate phylogenetic regression to explore the relationship between head shape and size. Is there a relationship between the two traits?
4. Assess the phylogenetic signal of head shape, dies head shape associate with the phylogeny?

</div>
<br></br>
